---
title: "Marginal survival probabilities from {brms} weibull models"
output: github_document
---

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}

knitr::opts_chunk$set(echo = TRUE,warning=FALSE, tidy = TRUE,message = FALSE, fig.dim = c(9,6))

```

```{r libraries}
library(tidyverse)
library(survival)
library(brms)
library(tidybayes)
library(here)

theme_set(theme_tidybayes())

```

We will define som usefull functions for working with the weibull distribution, first median survival as

```{r median weibull}
median_weibull  <-  function(scale, shape) {
    (log(2) ^ (1 / shape)) * scale}
```

*brms* uses a `scale,shape` parametrisation for the Weibull distribution but the main parameter in the model is $\mu$. We convert $\mu$ and `shape` to `scale` through as described [here](https://stats.stackexchange.com/questions/542106/difference-in-fitting-to-right-censored-data-between-mle-and-bayesian-method), using the formula $scale = \frac{\mu}{\Gamma(1+\frac{1}{shape})}$

```{r convert from mu and shape to scale}

weibull_mu_to_scale <-   function(mu, shape) {    mu / gamma(1 + 1 / shape)  }
```

We also need the Weibull survival function to calculate and plot the survival curves

```{r survival function}
weibull_survival <-  function(scale,shape,time){exp(-(time / scale) ^ shape)}
```



We'll use the `colon` data from the `survival` package, focusing on the mortality endpoint and the effect of `rx`, marginalising over other covariates. 

```{r load data}

colon <- colon |>
  filter(etype == 2) |>
  mutate(
    censored = 1 - status,
    across(rx, as.factor),
    across(sex,  ~ factor(., labels = c("Female", "Male"))),
    across(c(obstruct, perfor, adhere), as.factor),
    across(differ,  ~ ordered(., labels = c("well", "moderate", "poor"))),
    across(extent,  ~ ordered(., labels = c("submucosa", "muscle", "serosa", "contiguous structures"))),
    across(surg,  ~ ordered(., labels = c("short", "long")))
  )

```

```{r simple weibull fit}
formula_simple <- bf(time+1|cens(censored) ~ rx + sex + age + nodes + mo(differ) + adhere,
                     family = weibull)


prior_simple <- prior(normal(0, 2.5), class = b)




weibull_simple <- brm(formula = formula_simple,
                     prior = prior_simple,
                     data = colon,
                     file = here("fits/weibull_simple.rds"),
                     file_refit = "never")

weibull_simple
```


```{r simple weibull fit conditional effects}
plot(conditional_effects(weibull_simple),ask=FALSE)

```

```{r posterior survival simple weibull}
posterior_survival_probability <- weibull_simple |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_simple),
    value = "mu",
    transform = TRUE,
    dpar = "shape"
  ) |>
  mutate(scale = mu / gamma(1 + 1 / shape)) |> 
  group_by(rx, .draw, ) |>
  summarise(shape = mean(shape), scale = mean(scale)) |>
  ungroup() |>
  expand(nesting(.draw, rx, scale,shape), time = modelr::seq_range(colon$time, n = 101)) |>
  mutate(S = weibull_survival(scale,shape,time))


```

```{r median survival simple weibull}
median_survival <- posterior_survival_probability |> 
  select(-c(time,S)) |> 
  distinct() |> 
  group_by(.draw,rx) |> 
  summarise(median_survival=median_weibull(scale,shape)) |> 
  group_by(rx) |> 
  median_hdi(median_survival)


posterior_survival_probability |> 
  ggplot() +
  aes(x = time, y = S, group = rx) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_segment(
    aes(
      x = median_survival,
      xend = median_survival,
      y = 0,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival
  ) +
  geom_segment(
    aes(
      x = 0,
      xend = median_survival,
      y = 0.5,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival
  ) +
  facet_grid(rows = vars(rx)) +  scale_fill_brewer(name = "Confidence level") +
theme_light() +
  theme(legend.position = "bottom")+
  labs(title = "Marginal adjusted surival probabilities by treatment group", subtitle = "(Population averaged)", caption = "Dashed lines denote estimated median survival")

```


```{r scale and shape weibull distributional fit}

formula_shape_scale <- bf(time+1|cens(censored) ~ rx + sex + age + nodes + mo(differ) + adhere,
                          shape ~ 0+rx,
                     family = weibull)

get_prior(formula_shape_scale,
          data=colon)

prior_shape_scale <- prior(normal(0, 2.5), class = b) +
  prior(normal(0,0.5), class = b, dpar = shape)




weibull_shape_scale <- brm(formula = formula_shape_scale,
                     prior = prior_shape_scale,
                     data = colon,
                     file = here("fits/weibull_shape_scale.rds"),
                     file_refit = "never")

prior_summary(weibull_shape_scale)

plot(conditional_effects(weibull_shape_scale),ask=FALSE)



```

```{r weibull shape scale posterior marginal survival }
posterior_survival_probability_ss <- weibull_shape_scale |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_shape_scale),
    value = "mu",
    transform = TRUE,
    dpar = "shape"
  ) |>
  mutate(scale = weibull_mu_to_scale(mu,shape)) |>
  group_by(rx, .draw, ) |>
  summarise(shape = mean(shape), scale = mean(scale)) |>
  ungroup() |>
  expand(nesting(.draw, rx, scale,shape), time = modelr::seq_range(colon$time, n = 101)) |>
  mutate(S = weibull_survival(scale,shape,time))

```

```{r scale shape predicted survival plot}
posterior_survival_probability_ss |> 
  ggplot() +
  aes(x = time, y = S, group = rx) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_segment(
    aes(
      x = median_survival,
      xend = median_survival,
      y = 0,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival
  ) +
  geom_segment(
    aes(
      x = 0,
      xend = median_survival,
      y = 0.5,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival
  ) +
  facet_grid(rows = vars(rx)) +  scale_fill_brewer(name = "Confidence level") +
theme_light() +
  theme(legend.position = "bottom") +
  labs(title = "Marginal adjusted surival probabilities by treatment group", subtitle = "(Population averaged)", caption = "Dashed lines denote estimated median survival")

```

```{r compare rx parameters}
weibull_simple |>
  gather_draws(`b_rx.*`, regex = TRUE) |>
  mutate(model = "simple") |>
  bind_rows(
    weibull_shape_scale |>
      gather_draws(`b_rx.*`, regex = TRUE) |>
      mutate(model = "scale_shape")
  ) |> 
  ggplot() + 
  aes(x=.value,y=.variable,group=model, fill = model)+
  stat_gradientinterval(position = position_dodge(width = 1))
```

```{r compare marginal median survival estimates}
posterior_survival_probability |>
  select(-c(time, S)) |>
  distinct() |>
  group_by(.draw, rx) |>
  summarise(median_survival = median_weibull( scale, shape)) |>
  mutate(model = "simple") |>
  bind_rows(
    posterior_survival_probability_ss |>
      select(-c(time, S)) |>
      distinct() |>
      group_by(.draw, rx) |>
      summarise(median_survival = median_weibull(scale, shape)) |>
      mutate(model = "shape_scale")
  ) |>
  ggplot() +
  aes(x = median_survival, y = rx,group = model, fill =model ) +
  stat_gradientinterval(position = position_dodge(width = 1))+
  labs(title)
```
