---
title: "Marginal survival probabilities, marginal median survival and hazard ratios from {brms} Weibull models"
author: "Lars Mølgaard Saxhaug"
output: github_document
bibliography: references.bib
csl: jama.csl
---

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}


knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  tidy = TRUE,
  message = FALSE,
  fig.dim = c(9, 6)
)

```

```{r libraries}
library(tidyverse)
library(survival)
library(brms)
library(tidybayes)
library(here)

theme_set(theme_tidybayes())

```

This started out as an exercice in plotting survival curves from `brms`, including marginalising over baseline covariates, and expanded to trialing the posibilities that lies in estimating effects for both $scale$ and $shape$ in Weibull regression.

We will define som usefull functions for working with the Weibull distribution, first median survival as $$scale(log(2)^{(1/shape)}) $$

```{r median weibull}
median_weibull  <-  function(scale, shape) {
  (log(2) ^ (1 / shape)) * scale
}
```

`brms` uses a $scale,shape$ parametrisation for the Weibull distribution but the $shape$ is not estimated directly, rather $\mu$ is (which means exponentiating coefficients yields mean time ratios). We convert $\mu$ and $shape$ to $scale$ as described [here](https://stats.stackexchange.com/questions/542106/difference-in-fitting-to-right-censored-data-between-mle-and-bayesian-method), using the formula $$scale = \frac{\mu}{\Gamma(1+\frac{1}{shape})}$$

```{r convert from mu and shape to scale}
weibull_mu_to_scale <-   function(mu, shape) {
  mu / gamma(1 + 1 / shape)
}
```

We also need the Weibull survival function to calculate and plot the survival curves $$\displaystyle S(t)=e^{-(\frac{time}{scale})^{shape}}$$

```{r weibull_survival_function}
weibull_survival <-
  function(scale, shape, time) {
    exp(-(time / scale) ^ shape)
  }
```

Hazard function (for calculating hazard ratio) which is $$\displaystyle \mathrm{h}(scale;shape;time)=\frac{shape}{scale}{(\frac{time}{scale})}^{scale-1}$$

```{r weibull_hazard_function}
weibull_hazard <-
  function(scale, shape, time) {
    (shape / scale) * (time / scale) ^ (shape - 1)
  }
```

We'll use the `colon` data from the `survival` package, focusing on the mortality endpoint and the effect of `rx`, marginalising over other covariates.

```{r load_data}

colon <- colon |>
  filter(etype == 2) |>
  mutate(
    censored = 1 - status,# brms uses 0 as non-censored, 1 as right censored
    across(rx, as.factor),
    across(sex,  ~ factor(., labels = c("Female", "Male"))),
    across(c(obstruct, perfor, adhere), as.factor),
    across(differ,  ~ ordered(., labels = c("well", "moderate", "poor"))),
    across(extent,  ~ ordered(., labels = c("submucosa", "muscle", "serosa", "contiguous structures"))),
    across(surg,  ~ ordered(., labels = c("short", "long")))
  ) |>   drop_na(rx,sex,age,nodes,differ,adhere,extent)

```

### Proprtional weibull regression

By only allowing $mu$ (and thus $scale$ parameter) to wary among treatment groups, but with a constant common $shape$ parameter, we are estimating a proportionoal hazards model. $shape$ determines whether the hazards are decreasing when $shape<1$ ("infant mortality") or increasing ("aging") when $shape>1$.

> Ordinal vaiables are handled as monotonic effects @bürkner2020

```{r proprotional_weibull_fit}
formula_scale_only <-
  bf(time  |
       cens(censored) ~ rx + sex + age + nodes + mo(differ) + adhere + mo(extent),
     family = weibull)


prior_scale_only <- prior(normal(0, 2.5), class = b) 


weibull_scale_only <- brm(formula = formula_scale_only,
                     prior = prior_scale_only,
                     data = colon,
                     file = here("fits/weibull_scale_only.rds"),
                     file_refit = "never")

weibull_scale_only
```

We then estimate the expected the survival probability and hazard for each treatment group at each time point, marginalising over the population in the style of the `comparisons` function from the `marginaleffects` package. (Thanks to Andrea Discacciati for correcting erronous averaging of scale/shape rather than the survival probabilities themselves as reported in [@brilleman2020])

```{r posterior ph weibull}
proportional_posterior <- weibull_scale_only |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_scale_only), # duplicate the dataset for each treatment level
    value = "mu",
    transform = TRUE,
    dpar = "shape", # return shape in addition to mu
    ndraws = 1000 , 
    seed = 123
  ) |>
  mutate(scale = weibull_mu_to_scale(mu,shape)) |> # reconvert mu and shape to scale
  ungroup() |> 
  expand(nesting(.draw, rx, scale, shape),
         time = modelr::seq_range(colon$time, n = 101)) |>
  mutate(S = weibull_survival(scale, shape, time),
         # population averaged survival probability for each time point and rx group
         h = weibull_hazard(scale, shape, time)) |> # population averaged hazard for each time point and rx group
  group_by(time, .draw, rx) |> # group by .draw, time and rx to find the population averaged survival conditional on rx and time 
  summarise(S = mean(S), h = mean(h)) |> 
  ungroup()


```

Next we compute population averaged median survival times for each group.

```{r median_survival_prop_weibull}
median_survival_prop <- weibull_scale_only |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_scale_only),
    value = "mu",
    transform = TRUE,
    dpar = "shape",
    ndraws = 1000 ,
    seed = 123
  ) |>
  mutate(scale = weibull_mu_to_scale(mu, shape)) |>
  group_by(.draw, rx) |>
  summarise(median_survival = mean(median_weibull(scale, shape))) |>
  group_by(rx) |>
  median_hdi(median_survival)
```

Now we can plot population averaged survival curves as well as estimated median survival for each treatment group

```{r survival  plot proportional weibull}
proportional_posterior |> 
  ggplot() +
  aes(x = time, y = S, group = rx) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_segment(
    aes(
      x = median_survival,
      xend = median_survival,
      y = 0,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival_prop
  ) +
  geom_segment(
    aes(
      x = 0,
      xend = median_survival,
      y = 0.5,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival_prop
  ) +
  scale_y_continuous(labels = scales::percent_format())+
  facet_grid(rows = vars(rx)) +  scale_fill_brewer(name = "Confidence level") +
theme_light() +
  theme(legend.position = "bottom")+
  labs(title = "Adjusted surival probabilities by treatment group\nProportional Weibull model", subtitle = "(Population averaged)", caption = "Dashed lines denote estimated median survival")

```

Let's check that the hazard ratio is indeed constant (proportional):

```{r prop_haz_plot}
proportional_posterior |>
  group_by(rx, time) |>
  compare_levels(h,
                 by = rx,
                 fun = `/`,
                 comparison = "control") |>
  rename("HR" = h) |>
  ggplot() +
  aes(x = time, y = HR) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer() +
  facet_grid(rows = vars(rx))
```

### Non-proportional hazards Weibull modell

If we not only let $mu$ (and $scale$) vary, but also $shape$, we get a non-proportional hazard model, with a non-constant hazard ratio[@zuehlke2013].

```{r scale and shape weibull distributional fit}

formula_shape_scale <-
  bf( time | cens(censored) ~ rx + sex + age + nodes + mo(differ) + adhere + mo(extent),
    shape ~ 0 + rx,
    family = weibull
  )


prior_shape_scale <- prior(normal(0, 2.5), class = b) +
  prior(normal(0,0.5), class = b, dpar = shape) 




weibull_shape_scale <- brm(formula = formula_shape_scale,
                     prior = prior_shape_scale,
                     data = colon,
                     file = here("fits/weibull_shape_scale.rds"),
                     file_refit = "never")

prior_summary(weibull_shape_scale)

weibull_shape_scale



```

Repeating the above posterior calculations to arrvie at population averaged parameters, as well as time varying hazards and survival probablitities

```{r weibull shape scale posterior marginal survival }
nonprop_posterior <- weibull_shape_scale |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_shape_scale),
    value = "mu",
    transform = TRUE,
    dpar = "shape", # return shape in addition to mu
    ndraws = 1000 , 
    seed = 123
  ) |>
  mutate(scale = weibull_mu_to_scale(mu,shape)) |> # reconvert mu and shape to scale
  ungroup() |> 
  expand(nesting(.draw, rx, scale, shape),
         time = modelr::seq_range(colon$time, n = 101)) |>
  mutate(S = weibull_survival(scale, shape, time),         # population averaged survival probability for each time point and rx group
         h = weibull_hazard(scale, shape, time)) |> # population averaged hazard for each time point and rx group
  group_by(time, .draw, rx) |> # group by .draw, time and rx to find the population averaged survival conditional on rx and time 
  summarise(S = mean(S), h = mean(h)) |> 
  ungroup()

```

```{r nonprop median survival}
  median_survival_nonprop <- weibull_shape_scale |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_shape_scale),
    value = "mu",
    transform = TRUE,
    dpar = "shape", # return shape in addition to mu
    ndraws = 1000 , 
    seed = 123
  ) |> 
  mutate(scale = weibull_mu_to_scale(mu,shape)) |>
  group_by(.draw, rx) |>
  summarise(median_survival = mean(median_weibull(scale, shape))) |>
  group_by(rx) |>
  median_hdi(median_survival)
```

First let's plot the population averaged survival curves and estimated median survival estimates

```{r scale shape predicted survival plot}
nonprop_posterior |> 
  ggplot() +
  aes(x = time, y = S, group = rx) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_segment(
    aes(
      x = median_survival,
      xend = median_survival,
      y = 0,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival_nonprop
  ) +
  geom_segment(
    aes(
      x = 0,
      xend = median_survival,
      y = 0.5,
      yend = 0.5
    ),
    linetype = "dashed",
    data = median_survival_nonprop
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  facet_grid(rows = vars(rx)) +  scale_fill_brewer(name = "Confidence level") +
theme_light() +
  theme(legend.position = "bottom") +
  labs(title = "Adjusted surival curves by treatment group", subtitle = "(Population averaged)", caption = "Dashed lines denote estimated median survival")

```

Now let's plot the hazard ratios across time, demonstrating the non-proporitonal hazards

```{r compare rx HR}
nonprop_posterior |>
  group_by(rx, time) |>
  compare_levels(h,
                 by = rx,
                 fun = `/`,
                 comparison = "control") |>
  rename("HR" = h) |>
  ggplot() +
  aes(x = time, y = HR) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer() +
  facet_grid(rows = vars(rx)) +
  coord_cartesian(ylim=c(0,3))
```

And the population averaged hazard functions for each groups

```{r rx hazards}
nonprop_posterior |>
  group_by(rx, time) |>
  ggplot() +
  aes(x = time, y = h) +
  stat_lineribbon(.width = c(.99, .95, .8, .5), color = "#08519C") +
  scale_fill_brewer() +
  facet_grid(rows = vars(rx))


```

Finally, let's compare the population averaged median survival by group and model

```{r compare marginal median survival estimates}
weibull_shape_scale |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_scale_only),
    value = "mu",
    transform = TRUE,
    dpar = "shape", # return shape in addition to mu
    ndraws = 1000 , 
    seed = 123
  ) |> 
  mutate(scale = weibull_mu_to_scale(mu,shape)) |>
  group_by(.draw, rx) |>
  summarise(median_survival = mean(median_weibull(scale, shape)))|>
  mutate(Model = "PH Weibull") |>
  bind_rows(
    weibull_shape_scale |>
  linpred_draws(
    marginaleffects::datagridcf(rx = levels(colon$rx), model = weibull_shape_scale),
    value = "mu",
    transform = TRUE,
    dpar = "shape", # return shape in addition to mu
    ndraws = 1000 , 
    seed = 123
  ) |> 
  mutate(scale = weibull_mu_to_scale(mu,shape)) |>
  group_by(.draw, rx) |>
  summarise(median_survival = mean(median_weibull(scale, shape))) |>
      mutate(Model = "Non-PH Weibull")
  ) |>
  ggplot() +
  aes(x = median_survival,
      y = rx,
      group = Model,
      fill = Model) +
  stat_gradientinterval(position = position_dodge(width = 1)) +
  scale_x_continuous(name = "Median survival time") +
  labs(title = "Population averaged median survival by model and group")
```
